Large-scale text pre-training helps with dialogue act recognition, but not without fine-tuning

### 2021-03-12 reports/IWCS2021/iwcs-short-paper.tex
some cuts

Recently, large-scale neural language models, trained on massive corpora of text data, 
have achieved state-of-the-art results on a variety of traditional NLP tasks.
Given that dialogue, especially spoken dialogue, 
is radically different from the kind of data these language models are pre-trained on,
it is uncertain whether these models would be useful for dialogue-oriented tasks.
In the example from the Switchboard corpus, shown in Table~\ref{tab:example}, 
it is evident that the structure of dialogue is quite different from that of written text.
%even if one imagines a script for a movie or a play.
Not only is the internal structure of the contributions different---%
with features such as disfluencies, repair, incomplete sentences, and various vocal sounds---%
but the sequential structure of the discourse is different as well.
In dialogue, speakers take turns and switch perspectives.
More generally, the way that utterances contribute to and cohere with the discourse
is different from the relationship between sentences in a piece of written text.

@@

\paragraph{Fine-tuning vs.~further in-domain pre-training}
Transfer learning schemes for pre-trained language models such as BERT draw on one or more of the following three adaptation strategies \citep{sunHowFineTuneBERT2019}:
\emph{further pre-training}, in which the model is trained in an un-supervised way, similar to its initial training scheme, but on data that is in-domain for the target task; 
\emph{single-task fine-tuning}, in which the model's encoder layers are optimized during training for the target task;
and \emph{multi-task fine-tuning}, in which models with shared encoder layers are jointly trained on several related tasks.
In this paper, we investigate the first two methods and their combination, as applied to DAR.

Regardless of whether an encoder model has undergone further in-domain pre-training, the choice of whether to fine-tune during task training remains.
Freezing the encoder model during task training is more efficient, since the gradient of the loss function need only be computed for the task-specific layers.
However, fine-tuning can lead to better performance since the encoding itself is adapted to the target task and domain.

@@

\subsection{Pre-training vs. fine-tuning} \label{sec:experiment2} % Bill
First, we analyze how pre-training affects BERT's performance as an utterance encoder.
To do so, we consider the performance of DAR models with three different utterance encoders:
\begin{itemize}
  \item \texttt{BERT-FT} -- pre-trained BERT with DAR fine-tuning 
  \item \texttt{BERT-RI} -- randomly initialized BERT (with DAR fine-tuning)
  \item \texttt{BERT-FZ} -- pre-trained BERT without fine-tuning (frozen during DAR training)
\end{itemize}

\begin{table}[]
\centering
\begin{tabular}{@{}lrrrr@{}}
\toprule
                           & \multicolumn{2}{c}{SWDA}                          & \multicolumn{2}{c}{AMI-DA}                        \\ \midrule
                           & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} \\
\texttt{BERT-FT} & 36.75           & 76.60                    & 43.42                  & 64.93                    \\
\texttt{BERT-RI} & 32.18           & 73.80                    & 34.88                  & 60.89                    \\
\texttt{BERT-FZ} & 7.75            & 55.61                    & 14.86                  & 48.34 \\ \midrule
    Majority class    & 0.78  & 33.56 &  1.88 & 28.27      \\ \bottomrule
\end{tabular}
  \caption{Macro-average F1 and accuracy of BERT with standard pre-training and DAR fine-tuning (\texttt{BERT-FT}) vs.~the same model without pre-training (\texttt{BERT-RI}) and without fine-tuning (\texttt{BERT-FZ}).}
  \label{tab:exp2}
\end{table}

The pre-trained model is more accurate than the randomly initialized model by several percentage points on both DA corpora,
suggesting that BERT's extensive pre-training does provide some useful information for recognizing different dialogue acts (Table~\ref{tab:exp2}).
This performance boost is much more pronounced in the macro-averaged F1 score,
which is explained by the fact that at the tag level, pre-training has a larger impact on less frequent tags.
For a disaggregated view, see Figure 1 in the supplementary materials.

@@

Comparing the performance of the fine-tuned encoder to that of the frozen model, 
we see that pre-training alone is not enough.
The frozen model is heavily biased towards the most frequent tags, which explains its especially poor macro-average score (Table~\ref{tab:exp2}).
In SWDA, for example, the model with a frozen encoder predicts one of the two most common tags (Statement-non-opinion or Acknowledge) 86\% of the time, whereas those two
tags account for only 51\% of the ground truth tags.
The fine-tuned model is much less biased; it predicts the two most common tags only 59\% of the time.

@@

\subsection{Experiment 3: Impact of dialogue pre-training} \label{sec:experiment3} % Bill

Next, we assess the effect of additional dialogue pre-training on BERT's performance as an utterance encoder.\footnote{
In-domain pre-training is sometimes referred to as \textit{fine-tuning}, but we reserve that term for task-specific training on labeled data.}
\citet{sunHowFineTuneBERT2019} has reported that performing additional pre-training on unlabeled in-domain data improves performance on classification tasks. 
We want to see if BERT can benefit from pre-training, on dialogue data, including from data outside the immediate target domain.

@@

\begin{table}[]
\begin{tabular}{@{}lrrrr@{}}
\toprule
           & \multicolumn{2}{c}{SWDA}                          & \multicolumn{2}{c}{AMI-DA}                        \\ \midrule
           & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} \\
\texttt{BERT-FT}    & 36.75                  & 76.60                    & 43.42                  & 64.93                    \\
\texttt{BERT+ID-FT} & 43.63                  & 77.01                    & 46.70                  & \textbf{68.88}           \\
\texttt{BERT+CC-FT} & \textbf{47.78}         & \textbf{77.35}           & \textbf{48.86}         & 68.79                    \\ 
\texttt{BERT+OS-FT} & 41.42                  & 76.95                    & 48.65                  & 68.07                    \\ \midrule
\texttt{BERT-FZ}    &  7.75                  & 55.61           & 14.86                  & 48.34                    \\
\texttt{BERT+ID-FZ} &  6.46                  & 52.30                    & 14.48                  & 48.18                    \\
\texttt{BERT+CC-FZ} &  5.76                  & 51.14                    & 11.34                  & 40.48                    \\ 
\texttt{BERT+OS-FZ} &  \textbf{9.60}         & \textbf{57.67}           & \textbf{17.03}         & \textbf{51.03}           \\ \midrule
    Majority class  & 0.78                   & 33.56                    &  1.88                  & 28.27      \\ \bottomrule
\end{tabular}
\caption{Comparison of macro-average F1 and accuracy with further in-domain (\texttt{ID}), combined corpus (\texttt{CC}), and OpenSubtitles (\texttt{OS}) dialogue pre-training,
    for the frozen (\texttt{FZ}) and fine-tuned (\texttt{FT}) conditions.}
  \label{tab:exp3}
\end{table}

>>>>

Large-scale neural language models trained on massive corpora of text data
have achieved state-of-the-art results on a variety of traditional NLP tasks.
Given that dialogue, especially spoken dialogue, 
is radically different from the kind of data these language models are pre-trained on,
it is uncertain whether they would be useful for dialogue-oriented tasks.
In the example from the Switchboard corpus, shown in Table~\ref{tab:example}, 
it is evident that the structure of dialogue is quite different from that of written text.
%even if one imagines a script for a movie or a play.
Not only is the internal structure of contributions different---%
with features such as disfluencies, repair, incomplete sentences, and various vocal sounds---%
but the sequential structure of the discourse is different as well.
%In dialogue, speakers take turns and switch perspectives.
%More generally, the way that utterances contribute to and cohere with the discourse
%is different from the relationship between sentences in a piece of written text.

@@

\paragraph{Fine-tuning vs.~further in-domain pre-training}
We experiment with the following two transfer learning strategies \citep{sunHowFineTuneBERT2019}:
\emph{further pre-training}, in which the model is trained in an un-supervised way, similar to its initial training scheme, but on data that is in-domain for the target task; and
\emph{single-task fine-tuning}, in which the model's encoder layers are optimized during training for the target task.

Whether or not encoder model has undergone further in-domain pre-training, 
the choice of whether to fine-tune during task training remains.
Freezing the encoder model during task training is more efficient, 
since the gradient of the loss function need only be computed for the task-specific layers.
However, fine-tuning can lead to better performance since the encoding itself is adapted to the target task and domain.

@@

\subsection{Pre-training vs. fine-tuning} \label{sec:experiment2} % Bill
First, we analyze how pre-training affects BERT's performance as an utterance encoder.
To do so, we consider the performance of DAR models with three different utterance encoders:
\begin{itemize}
  \item \texttt{BERT-FT} -- pre-trained + DAR fine-tuning 
  \item \texttt{BERT-FZ} -- pre-trained, frozen during DAR
  \item \texttt{BERT-RI} -- randomly initialized + DAR fine-tuning 
\end{itemize}

%\begin{table}[]
%\centering
%\begin{tabular}{@{}lrrrr@{}}
%\toprule
                           %& \multicolumn{2}{c}{SWDA}                          & \multicolumn{2}{c}{AMI-DA}                        \\ \midrule
                           %& \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} \\
%\texttt{BERT-FT} & 36.75           & 76.60                    & 43.42                  & 64.93                    \\
%\texttt{BERT-RI} & 32.18           & 73.80                    & 34.88                  & 60.89                    \\
%\texttt{BERT-FZ} & 7.75            & 55.61                    & 14.86                  & 48.34 \\ \midrule
    %Majority class    & 0.78  & 33.56 &  1.88 & 28.27      \\ \bottomrule
%\end{tabular}
  %\caption{Macro-average F1 and accuracy of BERT with standard pre-training and DAR fine-tuning (\texttt{BERT-FT}) vs.~the same model without pre-training (\texttt{BERT-RI}) and without fine-tuning (\texttt{BERT-FZ}).}
  %\label{tab:exp2}
%\end{table}

The pre-trained model is more accurate than the randomly initialized model by several percentage points on both DA corpora,
suggesting that BERT's extensive pre-training does provide some useful information for recognizing different dialogue acts (Table~\ref{tab:results}).
This performance boost is much more pronounced in the macro-averaged F1 score,
which is explained by the fact that at the tag level, pre-training has a larger impact on less frequent tags.
For a disaggregated view, see Figure 1 in the supplementary materials.

@@

Comparing the performance of the fine-tuned encoder to that of the frozen model, 
we see that pre-training alone is not enough.
The frozen model is heavily biased towards the most frequent tags, which explains its especially poor macro-average score (Table~\ref{tab:results}).
In SWDA, for example, the model with a frozen encoder predicts one of the two most common tags (Statement-non-opinion or Acknowledge) 86\% of the time, whereas those two
tags account for only 51\% of the ground truth tags.
The model with a fine-tuned encoder is much less biased; it predicts the two most common tags only 59\% of the time.

@@

\subsection{Impact of dialogue pre-training} \label{sec:experiment3} % Bill

Next, we assess the effect of additional dialogue pre-training on BERT's performance as an utterance encoder.\footnote{
In-domain pre-training is sometimes referred to as \textit{fine-tuning}, but we reserve that term for task-specific training on labeled data.}
\citet{sunHowFineTuneBERT2019} has reported that performing additional pre-training on unlabeled in-domain data improves performance on classification tasks. 
We want to see if BERT can benefit from pre-training on dialogue data, including from data outside the immediate target domain.

@@

\begin{table}[]
\begin{tabular}{@{}lrrrr@{}}
\toprule
           & \multicolumn{2}{c}{SWDA}                          & \multicolumn{2}{c}{AMI-DA}                        \\ \midrule
           & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} & \multicolumn{1}{c}{F1} & \multicolumn{1}{c}{acc.} \\
\texttt{BERT-FT}    & 36.75                  & 76.60                    & 43.42                  & 64.93                    \\
\texttt{BERT+ID-FT} & 43.63                  & 77.01                    & 46.70                  & \textbf{68.88}           \\
\texttt{BERT+CC-FT} & \textbf{47.78}         & \textbf{77.35}           & \textbf{48.86}         & 68.79                    \\ 
\texttt{BERT+OS-FT} & 41.42                  & 76.95                    & 48.65                  & 68.07                    \\ \midrule
\texttt{BERT-FZ}    &  7.75                  & 55.61           & 14.86                  & 48.34                    \\
\texttt{BERT+ID-FZ} &  6.46                  & 52.30                    & 14.48                  & 48.18                    \\
\texttt{BERT+CC-FZ} &  5.76                  & 51.14                    & 11.34                  & 40.48                    \\ 
\texttt{BERT+OS-FZ} &  \textbf{9.60}         & \textbf{57.67}           & \textbf{17.03}         & \textbf{51.03}           \\ \midrule
\texttt{BERT-RI} & 32.18           & 73.80                    & 34.88                  & 60.89                    \\ \midrule
    Majority class  & 0.78                   & 33.56                    &  1.88                  & 28.27      \\ \bottomrule
\end{tabular}
\caption{Comparison of macro-average F1 and accuracy with further in-domain (\texttt{ID}), combined corpus (\texttt{CC}), and OpenSubtitles (\texttt{OS}) dialogue pre-training,
for the frozen (\texttt{FZ}) and fine-tuned (\texttt{FT}) conditions. \texttt{BERT-RI} uses a randomly initialized utterance encoder with no pre-training but with fine-tuning.}
  \label{tab:results}
\end{table}
